---
title: "Exmaple Usage"
author: "Zehui Yin"
date: "`r Sys.Date()`"
output: html_document
---

# Vignette

```{r setup}
library(sifr)
library(mapview)
```

In this example, I provided some example usage for the `sifr` package. Some 
sociodemographic and built environment variables are calculated for a small 
random sample of schools in Toronto.

# All types of School Locations in Toronto

Load data

```{r}
# school locations in Toronto
schools <- st_read("./../extdata/toronto_schools.gpkg")
# to force the geometry column to name geometry
st_geometry(schools) <- "geometry"

# for reproductivity
set.seed(123)

# take a random 30 sample to simplify calculation as this is just a example
schools <- schools[sample(1:nrow(schools), 30),]

mapview(schools)
```

```{r}
# road intersections in Toronto
intersections <- st_read("./../extdata/centreline_intersection.gpkg")
# to force the geometry column to name geometry
st_geometry(intersections) <- "geometry"

# road centrelines in Toronto
roads <- st_read("./../extdata/centreline.gpkg")
# to force the geometry column to name geometry
st_geometry(roads) <- "geometry"

# 2021 Canadian census data
census_data <- readRDS("./../data/census_data.rds")
```

# Calculate Number of Intersections within different buffers

```{r}
# generate what intersection falls within 300 meter buffers
school_int_sec <- what_within_each_stops(schools, intersections, 26918, 300)

# create holder to record how many intersections falls within
school_int_sec$input_rows_within_count <- 0

for (i in 1:nrow(school_int_sec)) { # loop over stop rows
  if(!is.na(str_to_num(i,"input_rows_within",school_int_sec))[1]){ # check whether there is intersection falls within
    school_int_sec$input_rows_within_count[i] <- length(str_to_num(i,"input_rows_within",school_int_sec)) # record the number of intersections
  }
}

schools$Rd_conn <- school_int_sec$input_rows_within_count
mapview(schools, cex = "Rd_conn")
```

# Calculate Length of Roads

```{r}
# generate what intersection falls within 300 meter buffers
road_length <- length_in_buffer(schools, roads, 26918, 300)
schools$road_len <- road_length$total_length_within_buffer

mapview(schools, cex = "road_len")
```

# Calculate Average Population Density

```{r}
# get rid of the long name
census_data$pop_den <- census_data$`v_CA21_6: Population density per square kilometre`

pop_den <- average_value_in_buffer(schools, census_data, 26918, 300, 
                                   "pop_den", FALSE)

schools$pop_den <- pop_den$pop_den

mapview(schools, cex = "pop_den")
```

# Assign the nearest Census Tract's Population Density (for median values)

```{r}
pop_den_median <- nearest_median_value(schools, census_data, "pop_den", 26918)

schools$pop_den_median <- pop_den$pop_den

mapview(schools, cex = "pop_den_median")
```

```{r eval=FALSE, include=FALSE}
rmarkdown::render("./example.Rmd")
```
